other functionalities

### 2025-02-27 inst/gitbook/05_other-functionalities.Rmd
Vignette update. Read to be pushed

This function allows to run ancestral characters estimations on multiple trees.
In it's most basic structure (e.g. using all default arguments) this function is using a mix of `ape::ace` and `castor::asr_mk_model` depending on the data and the situation and is generally faster than both functions when applied to a list of trees.
However, this function provides also some more complex and modular functionalities, especially appropriate when using discrete morphological character data.

### Using different character tokens in different situations

@@

>>>>

This function allows to run ancestral characters estimations on multiple trees.
In it's most basic structure (e.g. using all default arguments) this function is using a mix of `ape::ace` and `castor::asr_mk_model` depending on the data and the situation.
If you're running only a single trait estimation, using these functions is more recommended rather that using `multi.ace`.
However, this provides some more complex and modular functionalities which can make it more appropriate to use that their parent functions in the following cases:

 * Inferring multiple traits from a `"matrix"` or a `"data.frame"` (these all continuous or discrete traits, or a mix of both!);
 * Inferring traits using multiple trees;
 * Using traits estimations distributions (samples) rather than point estimates;
 * Inferring complex discrete characters;
 * Running parallel estimations;

The following tutorials cover some of the more advanced options in more details compared to the function manual (`?multi.ace`).

### Complex discrete characters: using different tokens in different situations

@@

### Getting sampled distributions instead of point estimates from the estimations

You can also use `multi.ace` to sample distributions instead of getting point estimates.
For example, by default, when using `multi.ace` on continuous characters, the output estimation is the average most likely value for the trait at every specific nodes.

```{r}
set.seed(1)
## Creating a coalescent tree
my_tree <- makeNodeLabel(rcoal(15))

## Creating a matrix of continuous characters
data <- space.maker(elements = 15, dimensions = 1, distribution = rnorm,
                    elements.name = my_tree$tip.label)

## Running the estimation with details
point_estimates <- multi.ace(data, my_tree, estimation.details = list("CI95"))
```

For example for trait 1 for Node 1, the estimated trait value is `r unname(round(point_estimates$estimations[1,1], 3))` but with a confidence interval of `r unname(round(point_estimates$details[[1]]$CI95[1,1], 3))` and `r unname(round(point_estimates$details[[1]]$CI95[1,2], 3))` which is quiet the spread.
To avoid treating the ancestral estimation as a point estimate (here the average of a large spread) we can also just get a number of samples to get an estimation of the distribution instead of the point estimate using the `sample` argument.

```{r}
## Getting 1000 samples from the distribution
distribution_estimates <- multi.ace(data, my_tree, sample = 1000)
```

This results in 1000 matrices that are sampled by default uniformaly.
We can visualise that for Node 1 that will be randomly sampled between the estimate's 95% confidence interval highlighted above.

```{r, fig.height = 4, fig.width = 4}
## Visualising the sampling of Node 1
node_1_estimates <- unlist(lapply(distribution_estimates, function(x) x[1,1]))
hist(node_1_estimates, main = "Distribution of estimates for Node 1")
# Note that the two small bars are because of the
# last and first bins should start/end at the actual 95 CI
```

Of course, we can change the type of function to use for sampling knowing the confidence interval of our estimations.
We can specify this by providing a named list containing the following two elements:

 1. `fun` that is the function to use for the sampling (`rnorm` in the example below);
 2. `param` that contains a list of functions or variables on how to calculate the extra arguments for `fun`. For example for `rnorm`, the function needs 1) a `mean` argument (here `mean = mean`, to use the function `mean` to calculate... the mean); and 2) a `sd` argument (here `sd = approx.sd`, a function to approximate the stnadard deviation knowing the confidence interval - assuming normality of the data).

```{r}
## Function for approximating the standard deviation
## from the range of a confidence interval.
approx.sd <- function(CI)return(diff(range(CI))/4)

## Getting 1000 samples from the distribution using a normal one
distribution_estimates <- multi.ace(data, my_tree, sample = 1000,
                            sample.fun = list(fun = rnorm,
                                              param = list(mean = mean,
                                                           sd = approx.sd)))
```

And again, we can visualise that:

```{r, fig.height = 4, fig.width = 4}
## Visualising the sampling of Node 1
node_1_estimates <- unlist(lapply(distribution_estimates, function(x) x[1,1]))
hist(node_1_estimates, main = "Distribution of estimates for Node 1")
# Which is now a neat normal distribution
```

In this example we used the default and a specified `sample.fun` argument for a matrix with only one trait.
However you can use any number of distributions for any number of traits: either just the one to be applied to all traits or a list of the same length as the number of traits with a list of `fun` and `param` for each.

#### Sampling discrete characters

You can also use the `sample` option on discrete characters.
In that case it's even easier because it doesn't need a `sample.fun` argument.
The character states estimates are just drawn from the scale likelihood estimate for each character (e.g. X% chance of being in state `0` and Y% chance of being in state `1` for a binary character).
